\changefontsize{1.12\defaultfontsize}

%\addtocounter{page}{-8}
%\origpage{xvii}
\headingthree{LIST OF SUBSCRIBERS.}
\markboth{list of subscribers.}{list of subscribers.}
\thispagestyle{empty}
\setcounter{page}{17}

\setlength{\columnseprule}{0.4pt}
\begin{multicols}{2}\small
\setlength{\parindent}{0em}
\setlength{\parskip}{0em}
\raggedright
\raggedcolumns
\subscriber{Ashe}, J. W. L., Esq., Exeter.

\subscriber{Allen}, G. F., Esq., Organist of St.
George’s, Wolverhampton.

\subscriber{Allen}, G. B., Esq., Armagh.

\subscriber{Alston}, The Rev. E. C., Dennington
Rectory.

\subscriber{Amott}, J., Esq., Organist of the Cathedral,
Gloucester.

\subscriber{Atkins}, R. A., Esq., Organist of the
Cathedral, St. Asaph’s.

\subscriber{Arcedeckne, Andrew}, Esq.

\subscriber{Ashburnham}, The Right Hon. the
Countess of.

\subscriber{Ashpitel, Arthur}, F.S.A.
\bigskip

\subscriber{Bartholomew}, Mrs. \textsc{Mounsey}.

\subscriber{Bacon}, G. P., Esq., Lewes.

\subscriber{Bagnold}, Mrs.

\subscriber{Barnett}, Mr. \textsc{John Francis}.

\subscriber{Barlow}, Mr. W., Manchester.

\subscriber{Beevor}, C., Esq.

\subscriber{Benedict, Jules}, Esq.

\subscriber{Best}, T. W., Esq., Organist of St. George’s
Hall, Liverpool.

\subscriber{Binfield}, Miss, Reading. 2 \textit{copies}.

\subscriber{Blake, Thomas}, Esq.

\subscriber{Blockley, John}, Esq.

\subscriber{Bloxsome, Charles}, Esq., Sheffield.

\subscriber{Bishop, John}, Esq., Cheltenham.

\subscriber{Barnett, John}, Esq., Cheltenham.

\subscriber{Blumenthal}, J., Esq.

\subscriber{Black}, Dr.

\subscriber{Bowers}, Miss, Clifton.

\subscriber{Bosanquet}, Mrs. \textsc{Godfrey}

\subscriber{Bourne}, Dr., Ashted House, Birmingham.

\subscriber{Braine}, W. R., Esq., Kensington.
\columnbreak

\subscriber{Broadwood, Henry} F., Esq.

\subscriber{Broadwood, Walter}, Esq.

\subscriber{Bros, Thomas}, Esq.

\subscriber{Brown, Graham}, Esq.

\subscriber{Brown}, T. F., Esq., Jersey.

\subscriber{Bruce}, Lord \textsc{Charles}.

\subscriber{Bucher}, M., Edinburgh.

\subscriber{Bussell}, H., Esq., Dublin.

\subscriber{Buck}, Dr., Norwich.

\subscriber{Bunnett}, E., Esq., Norwich.

\subscriber{Burchett}, J. R., Esq.

\subscriber{Benson}, G., Esq.

\subscriber{Birkenhead}, Miss.

\subscriber{Begrez}, J., Esq.

\subscriber{Butterworth}, Mr., Sheffield.
\bigskip

\subscriber{Cawdor}, The Right Hon. the Earl of.

\subscriber{Calthorp, Thomas}, D., Esq.

\subscriber{Cawood, Martin}, Esq., Leeds.

\subscriber{Callcott}, W. H., Esq., Kensington.

\subscriber{Chaplin}, Captain, Hastings.

\subscriber{Chater, George}, Esq.

\subscriber{Chisholm}, H. W., Esq.

\subscriber{Close, Thomas}, Esq., Nottingham.

\subscriber{Clarke}, Mr., Preston.

\subscriber{Cocks, Robert Lincoln}, Esq.

\subscriber{Cohen, Albert}, Esq.

\subscriber{Cooper}, G. \textsc{Armytage}, Esq., Lecturer on
Music.

\subscriber{Compton, Henry}, Esq., Melbourne, Australia.

\subscriber{Copeland}, W. R., Esq., Liverpool.

\subscriber{Cox}, Captain C. J., Fordwich House,
Kent.

\subscriber{Crotch}, Rev. R. W., Uphill House, near
Weston-super-Mare.
\columnbreak

\subscriber{Crossley}, G. J., Esq., Bowden, near Manchester.

\subscriber{Cummings}, — Esq.

\subscriber{Curtis}, W., Esq., Nottingham.
\bigskip

\subscriber{Davis}, Miss.

\subscriber{Davidson}, W., Esq., Aberdeen.

\subscriber{Davenport, John}, Esq.

\subscriber{De Capell Broke}, Lady.

\subscriber{De la Rue, William}, Esq. 2 \textit{copies}.

\subscriber{Delavante} and Co., Manchester.

\subscriber{De Monti}, Mr. J. H., Glasgow. 2 \textit{copies}.

\subscriber{De Vos}, Polydore, Esq., Jersey.

\subscriber{Dixon, Robert} W., Esq., Seaton Carew.

\subscriber{Done}, W., Esq., Organist of the Cathedral,
Worcester.

\subscriber{Doneraile}, The Right Hon. Viscountess.

\subscriber{Duck}, Mr. W., Bath.

\subscriber{Duerdin, John}, Esq. 2 \textit{copies}.

\subscriber{Dunraven}, The Right Hon. the Earl of.

\subscriber{Duggan, Joseph}, Esq.

\subscriber{Dyce}, The Rev. \textit{Alexander}.
\bigskip

\subscriber{Eborall}, Miss, Lichfield.

\subscriber{Edgar}, Mrs. \textsc{George}.

\subscriber{Ella, John}, Esq. 2 \textit{copies}.

\subscriber{Engel, Carl}, Esq.

\subscriber{Euing, William}, F.S.A., Scot., Glasgow.
\bigskip

\subscriber{Favarger, Réné}, Esq.

\subscriber{Fagg}, Mr., Hull.

\subscriber{Farren, William}, Esq.

\subscriber{Farrer}, T. H., Esq.

\subscriber{Fellowes}, Lady.

\subscriber{Ferrari}, Signor \textsc{Adolfo}.

\subscriber{Fitzwilliam}, Mrs. \textsc{Edward}.

\subscriber{Foster}, John, Esq., Gent, of Her Majesty’s
Chapels Royal.

\subscriber{Fowler, Charles}, Esq., Torqnay.
\bigskip

\subscriber{Grattan}, H. W., Esq.

\subscriber{Giraud, Frank}, Esq., Faversham.

\subscriber{Gibbs, Walter} S., Esq., Bath.

\subscriber{Green, John}, Esq.

\subscriber{Green}, Miss, Ashby de la Zouche.

\subscriber{Goss}, The Rev. \textsc{John}, Hereford.
\columnbreak

\subscriber{Goss, John}, Esq., Organist and Composer
to Her Majesty’s Chapels Royal.

\subscriber{Godfrey, Charles}, Esq.
\bigskip

\subscriber{Hutton, Maxwell}, Esq., Dublin.

\subscriber{Harrison}, J., Esq., Deal.

\subscriber{Hacking, Richard}, Esq., Mus. Bac., Bury,
Lancashire,

\subscriber{Harrison}, W., Esq., F.G.S., F.Z.S., Galligreaves
House.

\subscriber{Hale}, Mr. C., Cheltenham.

\subscriber{Hawkins}, The Rev. W. B.

\subscriber{Hayes}, Mrs., St. Catharine’s.

\subscriber{Harris}, Mr., Northampton.

\subscriber{Harvey}, The Rev. G. H., Wimbledon.

\subscriber{Hammond}, Mr.

\subscriber{Hayward, Henry}, Esq., Wolverhampton.

\subscriber{Hallé, Charles}, Esq.

\subscriber{Hatton}, J. L., Esq.

\subscriber{Hendrie}, Robert, Esq.

\subscriber{Hennell}, G. R., Esq.

\subscriber{Hemery}, Mr.

\subscriber{Hills, William}, Esq.

\subscriber{Higham}, F., Esq., Wolverhampton.

\subscriber{Hime}, Mr. B., Manchester.

\subscriber{Holl, William}, Esq.

\subscriber{Holl, Henry}, Esq.

\subscriber{Holl, Francis}, Esq.

\subscriber{Hopkins}, E. J., Esq., Organist of the
Temple Church.

\subscriber{Holmes}, W. H., Esq.

\subscriber{Hopkins}, J. L., Mus. Doc., Cambridge.

\subscriber{Howe}, H., Esq., Isleworth.

\subscriber{Hopkinson}, Messrs., Leeds.

\subscriber{Horne, F. Lennox}, Esq.

\subscriber{Holland}, W., Esq.

\subscriber{Hullah}, J. P., Esq.

\subscriber{Hutton}, The Rev. \textsc{Hugh}.
\bigskip

\subscriber{Isaac}, B. R., Esq., Liverpool.
\bigskip

\subscriber{Jackson}, Mr., Bradford.

\subscriber{Jewson}, J. P., Esq., Stockton-on-Tees.
\bigskip

\subscriber{Knapton}, Miss, Ripon.

\subscriber{Kerslake}, Mr., Bristol. 2 \textit{copies}.

\subscriber{Kreeft}, S. C., Esq., Consul-General.
\pagebreak

\subscriber{Kendall}, J. F., Esq.

\subscriber{Kerry}, John, Esq.

\subscriber{Keene, Charles}, Esq.

\subscriber{Ketelle}, S. W., Esq., Newcastle.

\subscriber{Kirk}, J. M., Esq., Halifax.

\subscriber{King, H. Stavely}, Esq., M.D.

\subscriber{Knight}, Mr., Chichester.

\subscriber{King, Donald}, Esq.
\bigskip

\subscriber{Lablache}, Signor F.

\subscriber{Land, Edward}, Esq.

\subscriber{Lambert}, D., Esq., York.

\subscriber{Lambeth, Henry} A., Esq., Glasgow.

\subscriber{Lewis}, Miss, Putney.

\subscriber{Linwood}, Miss.

\subscriber{Lohr}, G. A., Esq., Leicester.

\subscriber{Lover, Samuel}, Esq.

\subscriber{Lovett}, Mr.

\subscriber{Lockey, Charles}, Esq.
\bigskip

\subscriber{Mason, Joseph Edward}, Esq.

\subscriber{Mason, Thomas}, jun., Esq., Newcastle-under-Lyme.

\subscriber{Marshall}, J. W., Esq., Richmond, Yorkshire.

\subscriber{Masson}, Miss.

\subscriber{Martineau, Philip,} Esq.

\subscriber{Mackenzie, J. Whitefoord}, Esq., Edinburgh.

\subscriber{Martin}, G. W., Esq.

\subscriber{Mackway}, Mr.

\subscriber{Mercer}, The Rev. W., M.A., Sheffield.

\subscriber{Mills}, R., jun., Esq.

\subscriber{Morant}, The Lady \textsc{Henrietta}.

\subscriber{Morant, John}, Esq., Brockenhurst House,
Lymington.

\subscriber{Monk, Edwin George}, Mus. Doc., Organist
of the Cathedral, York.

\subscriber{Monk}, W. H., Esq., Organist of King’s
College, London.

\subscriber{Morris, Val}., Esq.

\subscriber{Murby, Thomas}, Esq.

\subscriber{Mario}, Signor.

\subscriber{Maclure, Andrew}, Esq.

\subscriber{Morier}, G. J., Esq.

\subscriber{Moore}, The Hon. Mrs. \textsc{Montgomery},
Dublin.
\columnbreak

\subscriber{Newman}, The Rev. W. A., D.D., Wolverhampton.

\subscriber{Nichol}, — Esq.

\subscriber{Nicholls}, W. H., Esq.

\subscriber{Nicholson, Alfred}, Esq.

\subscriber{Norbury}, The Right Hon. the Earl of.

\subscriber{Norman}, Rev. R. W., St. Peter’s College,
near Abingdon.

\subscriber{Norwood, Jasper}, Esq., Preston.

\subscriber{Norman}, Mrs., The Rookery, Bromley.
\bigskip

\subscriber{Ollivier}, Mr. \textsc{Robert}.

\subscriber{Osborne}, G. A., Esq.

\subscriber{Ouseley}, The Rev. Sir \textsc{Frederick Arthur}
\subscriber{Gore}, Bart., Professor of Music in the
University of Oxford.

\subscriber{Ould, Edwin}, Esq.
\bigskip

\subscriber{Paget}, R., Esq.

\subscriber{Parker, John William}, Esq.

\subscriber{Paterson}, Mr. R. \textsc{Roy}, Edinburgh.

\subscriber{Pole, S. Chandos}, Esq.

\subscriber{Purdie}, Mr. \textsc{John}, Edinburgh.

\subscriber{Pye, Kellow}, Esq., Wimbledon.
\bigskip

\subscriber{Rawlings}, — Esq., Shrewsbury.

\subscriber{Ravey}, Mr.

\subscriber{Reed, T. German}, Esq.

\subscriber{Richards}, Mr. T.

\subscriber{Richardson}, — Esq., Swindon.

\subscriber{Romer, Frank}, Esq.

\subscriber{Robinson, Joseph}, Esq., Dublin.

\subscriber{Rowe}, Mrs., Plymouth.

\subscriber{Rhodes, Jeremiah}, Esq., Pontefract.

\subscriber{Robb}, Mrs.

\subscriber{Robeck}, The Baron de, Swordlestown.

\subscriber{Rudall, Rose, and Carte}, Messrs.
\bigskip

\subscriber{Stanford, John}, Esq., Dublin.

\subscriber{Salaman, Charles}, Esq.

\subscriber{Sandys}, W., Esq., F.S.A.

\subscriber{Salt, George}, M., Esq., Shrewsbury.

Sacred Harmonic Society (The)

\subscriber{Spark}, W., Esq., Organist, Leeds.

\subscriber{Smith, G. Townshend}, Esq., Organist of the
Cathedral, Hereford.
\pagebreak

\subscriber{Spencer}, The Rev. \textsc{Peter}, Temple Ewell,
near Dover.

\subscriber{Simms}, E., Esq., Coventry.

\subscriber{Simms}, Mr. H., Bath.

\subscriber{Smart}, Sir \textsc{George}.

\subscriber{Smith, Montem}, Esq.

\subscriber{Smith, Albert}, Esq.

\subscriber{Smith, Samuel}, Esq., Bradford.

\subscriber{Smith, W. H}., Esq., Sheffield.

\subscriber{Smith}, Mr. \textsc{John}, Leeds.

\subscriber{Smith, George}, Esq.

\subscriber{Singleton}, Rev. R. C., Kingstown, Dublin.

Signet Library, (The) Edinburgh.

\subscriber{Silver, William}, Esq.

\subscriber{Swift, Joseph}, Esq.

\subscriber{Skelton}, G. J., Esq., Hull.

\subscriber{Shea, Alexande}r, Esq.

\subscriber{Shelmerdine}, W., Esq., Nottingham.

Shoreham College.

\subscriber{Stanyon, John}, Esq., Leicester.

\subscriber{Stanley}, E. H., Esq.

\subscriber{Stephenson}, W. F., Esq., Organist, Bishopton.

\subscriber{Sloper, Lindsay}, Esq.

\subscriber{Sutton} and \textsc{Potter}, Messrs., Dover.

\subscriber{Stocqueler}, J. H., Esq.

\subscriber{Sykes, Luke}, R., Esq.

\subscriber{Spalding}, S., Esq.

\subscriber{Squire}, F., Esq.

\subscriber{Stokes}, Dr., Dublin.
\bigskip

\subscriber{Travers}, Miss M.T., Hartsbourne, Bushey,
Herts.

\subscriber{Thacker}, A. C., Esq., Peterborough.

\subscriber{Taylor, Hussey}, Esq.

\subscriber{Taylor, Bianchi}, Esq., Bath.

\subscriber{Thackeray}, W., Esq.

\subscriber{Tait}, W., Esq., Melrose.
%\columnbreak

\subscriber{Thomson, James}, Esq., Glasgow.

\subscriber{Thrupp, John}, Esq.

\subscriber{Turner}, Mr., Stockport. 3 \textit{copies}.

\subscriber{Turle}, Miss, Lyme Regis.

\subscriber{Travers}, Miss, Hillingdon.
\bigskip

\subscriber{Vernon}, The Right. Hon. Lord.

\subscriber{Vantini, Townsend}, Esq.
\bigskip

\subscriber{Ward}, The Right Hon. Lord. 2 \textit{copies}.

\subscriber{Walker, George}, Esq., Aberdeen.

\subscriber{Wallerstein}, F. G., Newcastle-on-Tyne.

\subscriber{Walsh}, E., Esq.

\subscriber{Warren}, E., Esq., Organist of Christ
Church, Dover.

\subscriber{Wallace, W. Vincent}, Esq.

\subscriber{Ward}, The Rev. \textsc{Arthur} R., St. John’s
College, Cambridge.

\subscriber{Warren}. Mr. \textsc{John}, Royston.

\subscriber{Weiss}, W. H., Esq.

\subscriber{Werner, Louis}, Esq.

\subscriber{Webster, Alfred}, Esq., Bath.

\subscriber{Williams}, T., Esq., Fairlight, Tunbridge
Wells.

\subscriber{Wighton}, Mr. A. J., Dundee.

\subscriber{Wilson}, M. C., Esq.

\subscriber{Winslow, Forbes}, Esq., M.D., D.C.L.

\subscriber{Winn}, W., Esq.

\subscriber{Winstanley}, J. B., Esq., Bramston House,
near Leicester.

\subscriber{Woolcombe}, The Rev. W. \textsc{Walker}, Manchester.

\subscriber{Wood}, Messrs. J. M., and Co., Glasgow.
2~\textit{copies}.

\subscriber{Wood}, Messrs., Edinburgh. 2~\textit{copies}.

\subscriber{Woodman}, W., Esq., Hobhill, Morpeth.

\subscriber{Wylde, Henry}, Esq., Mus. Doc.
\end{multicols}
\setlength{\columnseprule}{0pt}


