\frontmatter

%===============================================================================

\thispagestyle{empty}
%\changefontsize{0.85\defaultfontsize}
\begin{center}

%\vspace*{\baselineskip}

\Huge POPULAR MUSIC
\titlespace

\small OF THE

\titlespace

\Huge OLDEN TIME:

\titlespace

\small A COLLECTION OF

\titlespace

\huge ANCIENT SONGS, BALLADS,

\titlespace

\small AND

\titlespace

\huge DANCE TUNES,

\titlespace

\small ILLUSTRATIVE OF THE

\titlespace

\huge NATIONAL MUSIC OF ENGLAND.

\titlespace

\normalsize WITH SHORT INTRODUCTIONS TO THE DIFFERENT REIGNS,

AND NOTICES OF THE AIRS FROM WRITERS OF THE

SIXTEENTH AND SEVENTEENTH CENTURIES.

\titlespace

\small ALSO

\titlespace

\large A SHORT ACCOUNT OF THE MINSTRELS.

\titlespace

\small BY

\titlespace

\Large W. CHAPPELL, F.S.A.

\titlespace

\normalsize THE WHOLE OF THE AIRS HARMONIZED BY G. A. MACFARREN.

\titlespace

\large VOL. II.

\titlespace

\footnotesize “Prout sunt illi Anglicani concentus suavissimi quidem, ac elegantes.”

\textit{Thesaurus Harmonicus} \textsc{Laurencini}, \textit{Romani}, 1603.

\titlespace

\small 1855 Edition LONDON:

\textsc{Cramer, Beale \& Chappell, 201 Regent Street}

\titlespace

2021 Edition MILTON KEYNES

\textsc{Catrah Press}

\normalsize
\vfill
\end{center}

\pagebreak

%===============================================================================



\vspace*{12\baselineskip}

\begin{center}
{\large{CONTENTS OF VOL. II.}}
\end{center}

\vspace*{2\baselineskip}

\centerrule

\noindent\textsc{\hfill page}

\tocstyle{Conjectures as to Robin Hood}{387}

\tocstyle{Ballads relating to the adventures of Robin Hood.}{391
to 400}

\tocstyle{Puritanism in its effects upon Music and its
accessories, and Introduction to the Commonwealth period.}{401 to
424}

\tocstyle{Songs and ballads of the civil war, and of the time of
Cromwell.}{425-466}

\tocstyle{Introduction to the reign of Charles II.}{467}

\tocstyle{Songs and ballads from Charles II. to William and
Mary.}{491 to 608}

\tocstyle{Remarks on Anglo-Scottish songs.}{609 to 616}

\tocstyle{Specimens of ditto.}{617 to 620}

\tocstyle{Introduction to the reigns of Queen Anne, George I.,
and George II.}{621 to 632}

\tocstyle{Songs and ballads of ditto.}{633 to 726}

\tocstyle{Traditional songs of uncertain date.}{727 to 749}

\tocstyle{Religious Christmas Carols.}{750 to 758}

\tocstyle{Appendix, consisting of additions to the Introductions,
and of further remarks upon the tunes included in both volumes.}{759
to 788}

\tocstyle{Characteristics of National English Airs, and
summary.}{789 to 797}

\vspace{2\baselineskip}
\begin{center}\rule{4em}{0.4pt}\end{center} \vspace*{4\baselineskip}
\vfill
%    \pagebreak


\intentionalemptypage \mainmatter \setcounter{page}{387}
\thispagestyle{empty} 


\vspace*{2\baselineskip}

\begin{center} \Large{POPULAR MUSIC}

\medskip

\normalsize\textsc{OF THE OLDEN TIME.} \end{center}


\centerrule

