%===============================================================================
\renewcommand\versoheader{list of subscribers.}
\renewcommand\rectoheader{list of subscribers.}
\changefontsize[1.4\defaultfontsize]{\defaultfontsize}


%xvii
\chapter{LIST OF SUBSCRIBERS.}
\thispagestyle{empty}
\setcounter{page}{17}

\setlength{\columnseprule}{0.4pt}
\begin{multicols}{2}\small
\raggedright
\subscriber{Ashe}, J. W. L., Esq., Exeter.

\subscriber{Allen}, G. F., Esq., Organist of St.~George’s, Wolverhampton.

\subscriber{Allen}, G. B., Esq., Armagh.

\subscriber{Alston}, The Rev. E. C., Dennington
Rectory.

\subscriber{Amott}, J., Esq., Organist of the Cathedral,
Gloucester.

\subscriber{Atkins}, R. A., Esq., Organist of the
Cathedral, St. Asaph’s.

\subscriber{Arcedeckne, Andrew}, Esq.

\subscriber{Ashburnham}, The Right Hon. the
Countess of.

\subscriber{Ashpitel, Arthur}, F.S.A.
\bigskip

\subscriber{Bartholomew}, Mrs. \textsc{Mounsey}.

\subscriber{Bacon}, G. P., Esq., Lewes.

\subscriber{Bagnold}, Mrs.

\subscriber{Barnett}, Mr. \textsc{John Francis}.

\subscriber{Barlow}, Mr. W., Manchester.

\subscriber{Beevor}, C., Esq.

\subscriber{Benedict, Jules}, Esq.

\subscriber{Best}, T. W., Esq., Organist of St.~George’s
Hall, Liverpool.

\subscriber{Binfield}, Miss, Reading. 2 \textit{copies}.

\subscriber{Blake, Thomas}, Esq.

\subscriber{Blockley, John}, Esq.

\subscriber{Bloxsome, Charles}, Esq., Sheffield.

\subscriber{Bishop, John}, Esq., Cheltenham.

\subscriber{Barnett, John}, Esq., Cheltenham.

\subscriber{Blumenthal}, J., Esq.

\subscriber{Black}, Dr.

\subscriber{Bowers}, Miss, Clifton.

\subscriber{Bosanquet}, Mrs. \textsc{Godfrey}

\subscriber{Bourne}, Dr., Ashted House, Birmingham.

\subscriber{Braine}, W. R., Esq., Kensington.
%\columnbreak

\subscriber{Broadwood, Henry} F., Esq.

\subscriber{Broadwood, Walter}, Esq.

\subscriber{Bros, Thomas}, Esq.

\subscriber{Brown, Graham}, Esq.

\subscriber{Brown}, T. F., Esq., Jersey.

\subscriber{Bruce}, Lord \textsc{Charles}.

\subscriber{Bucher}, M., Edinburgh.

\subscriber{Bussell}, H., Esq., Dublin.

\subscriber{Buck}, Dr., Norwich.

\subscriber{Bunnett}, E., Esq., Norwich.

\subscriber{Burchett}, J. R., Esq.

\subscriber{Benson}, G., Esq.

\subscriber{Birkenhead}, Miss.

\subscriber{Begrez}, J., Esq.

\subscriber{Butterworth}, Mr., Sheffield.
\bigskip

\subscriber{Cawdor}, The Right Hon. the Earl of.

\subscriber{Calthorp, Thomas}, D., Esq.

\subscriber{Cawood, Martin}, Esq., Leeds.

\subscriber{Callcott}, W. H., Esq., Kensington.

\subscriber{Chaplin}, Captain, Hastings.

\subscriber{Chater, George}, Esq.

\subscriber{Chisholm}, H. W., Esq.

\subscriber{Close, Thomas}, Esq., Nottingham.

\subscriber{Clarke}, Mr., Preston.

\subscriber{Cocks, Robert Lincoln}, Esq.

\subscriber{Cohen, Albert}, Esq.

\subscriber{Cooper}, G. \textsc{Armytage}, Esq., Lecturer on
Music.

\subscriber{Compton, Henry}, Esq., Melbourne, Australia.

\subscriber{Copeland}, W. R., Esq., Liverpool.

\subscriber{Cox}, Captain C. J., Fordwich House,
Kent.

\subscriber{Crotch}, Rev. R. W., Uphill House, near
Weston-super-Mare.
\end{multicols}
%xvii
\pagebreak