 

%xviii
\changefontsize[1.4\defaultfontsize]{1.00\defaultfontsize}
\begin{multicols}{2}\small
\setlength{\parindent}{0em}
\setlength{\parskip}{0em}
\raggedright

\subscriber{Crossley}, G. J., Esq., Bowden, near Manchester.

\subscriber{Cummings}, — Esq.

\subscriber{Curtis}, W., Esq., Nottingham.
\bigskip

\subscriber{Davis}, Miss.

\subscriber{Davidson}, W., Esq., Aberdeen.

\subscriber{Davenport, John}, Esq.

\subscriber{De Capell Broke}, Lady.

\subscriber{De la Rue, William}, Esq. 2 \textit{copies}.

\subscriber{Delavante} and Co., Manchester.

\subscriber{De Monti}, Mr. J. H., Glasgow. 2 \textit{copies}.

\subscriber{De Vos}, Polydore, Esq., Jersey.

\subscriber{Dixon, Robert} W., Esq., Seaton Carew.

\subscriber{Done}, W., Esq., Organist of the Cathedral,
Worcester.

\subscriber{Doneraile}, The Right Hon. Viscountess.

\subscriber{Duck}, Mr. W., Bath.

\subscriber{Duerdin, John}, Esq. 2 \textit{copies}.

\subscriber{Dunraven}, The Right Hon. the Earl of.

\subscriber{Duggan, Joseph}, Esq.

\subscriber{Dyce}, The Rev. \textit{Alexander}.
\bigskip

\subscriber{Eborall}, Miss, Lichfield.

\subscriber{Edgar}, Mrs. \textsc{George}.

\subscriber{Ella, John}, Esq. 2 \textit{copies}.

\subscriber{Engel, Carl}, Esq.

\subscriber{Euing, William}, F.S.A., Scot., Glasgow.
\bigskip

\subscriber{Favarger, Réné}, Esq.

\subscriber{Fagg}, Mr., Hull.

\subscriber{Farren, William}, Esq.

\subscriber{Farrer}, T. H., Esq.

\subscriber{Fellowes}, Lady.

\subscriber{Ferrari}, Signor \textsc{Adolfo}.

\subscriber{Fitzwilliam}, Mrs. \textsc{Edward}.

\subscriber{Foster}, John, Esq., Gent, of Her Majesty’s
Chapels Royal.

\subscriber{Fowler, Charles}, Esq., Torquay.
\bigskip

\subscriber{Grattan}, H. W., Esq.

\subscriber{Giraud, Frank}, Esq., Faversham.

\subscriber{Gibbs, Walter} S., Esq., Bath.

\subscriber{Green, John}, Esq.

\subscriber{Green}, Miss, Ashby de la Zouche.

\subscriber{Goss}, The Rev. \textsc{John}, Hereford.

\subscriber{Goss, John}, Esq., Organist and Composer
to Her Majesty’s Chapels Royal.

\subscriber{Godfrey, Charles}, Esq.
\bigskip

\subscriber{Hutton, Maxwell}, Esq., Dublin.

\subscriber{Harrison}, J., Esq., Deal.

\subscriber{Hacking, Richard}, Esq., Mus. Bac., Bury,
Lancashire,

\subscriber{Harrison}, W., Esq., F.G.S., F.Z.S., Galligreaves
House.

\subscriber{Hale}, Mr. C., Cheltenham.

\subscriber{Hawkins}, The Rev. W. B.

\subscriber{Hayes}, Mrs., St. Catharine’s.

\subscriber{Harris}, Mr., Northampton.

\subscriber{Harvey}, The Rev. G. H., Wimbledon.

\subscriber{Hammond}, Mr.

\subscriber{Hayward, Henry}, Esq., Wolverhampton.

\subscriber{Hallé, Charles}, Esq.

\subscriber{Hatton}, J. L., Esq.

\subscriber{Hendrie}, Robert, Esq.

\subscriber{Hennell}, G. R., Esq.

\subscriber{Hemery}, Mr.

\subscriber{Hills, William}, Esq.

\subscriber{Higham}, F., Esq., Wolverhampton.

\subscriber{Hime}, Mr. B., Manchester.

\subscriber{Holl, William}, Esq.

\subscriber{Holl, Henry}, Esq.

\subscriber{Holl, Francis}, Esq.

\subscriber{Hopkins}, E. J., Esq., Organist of the
Temple Church.\looseness=-1

\subscriber{Holmes}, W. H., Esq.

\subscriber{Hopkins}, J. L., Mus. Doc., Cambridge.

\subscriber{Howe}, H., Esq., Isleworth.

\subscriber{Hopkinson}, Messrs., Leeds.

\subscriber{Horne, F. Lennox}, Esq.

\subscriber{Holland}, W., Esq.

\subscriber{Hullah}, J. P., Esq.

\subscriber{Hutton}, The Rev. \textsc{Hugh}.
\bigskip

\subscriber{Isaac}, B. R., Esq., Liverpool.
\bigskip

\subscriber{Jackson}, Mr., Bradford.

\subscriber{Jewson}, J. P., Esq., Stockton-on-Tees.
\bigskip

\subscriber{Knapton}, Miss, Ripon.

\subscriber{Kerslake}, Mr., Bristol. 2 \textit{copies}.

\subscriber{Kreeft}, S. C., Esq., Consul-General.
\end{multicols}
%xviii
\pagebreak