%xix
\changefontsize[1.4\defaultfontsize]{1.00\defaultfontsize}
\begin{multicols}{2}\small
\setlength{\parindent}{0em}
\setlength{\parskip}{0em}
\raggedright
\raggedcolumns

\subscriber{Kendall}, J. F., Esq.

\subscriber{Kerry}, John, Esq.

\subscriber{Keene, Charles}, Esq.

\subscriber{Ketelle}, S. W., Esq., Newcastle.

\subscriber{Kirk}, J. M., Esq., Halifax.

\subscriber{King, H. Stavely}, Esq., M.D.

\subscriber{Knight}, Mr., Chichester.

\subscriber{King, Donald}, Esq.
\bigskip

\subscriber{Lablache}, Signor F.

\subscriber{Land, Edward}, Esq.

\subscriber{Lambert}, D., Esq., York.

\subscriber{Lambeth, Henry} A., Esq., Glasgow.

\subscriber{Lewis}, Miss, Putney.

\subscriber{Linwood}, Miss.

\subscriber{Lohr}, G. A., Esq., Leicester.

\subscriber{Lover, Samuel}, Esq.

\subscriber{Lovett}, Mr.

\subscriber{Lockey, Charles}, Esq.
\bigskip

\subscriber{Mason, Joseph Edward}, Esq.

\subscriber{Mason, Thomas}, jun., Esq., Newcastle-under-Lyme.

\subscriber{Marshall}, J. W., Esq., Richmond, Yorkshire.

\subscriber{Masson}, Miss.

\subscriber{Martineau, Philip,} Esq.

\subscriber{Mackenzie, J. Whitefoord}, Esq., Edinburgh.

\subscriber{Martin}, G. W., Esq.

\subscriber{Mackway}, Mr.

\subscriber{Mercer}, The Rev. W., M.A., Sheffield.

\subscriber{Mills}, R., jun., Esq.

\subscriber{Morant}, The Lady \textsc{Henrietta}.

\subscriber{Morant, John}, Esq., Brockenhurst House,
Lymington.

\subscriber{Monk, Edwin George}, Mus. Doc., Organist
of the Cathedral, York.

\subscriber{Monk}, W. H., Esq., Organist of King’s
College, London.

\subscriber{Morris, Val}., Esq.

\subscriber{Murby, Thomas}, Esq.

\subscriber{Mario}, Signor.

\subscriber{Maclure, Andrew}, Esq.

\subscriber{Morier}, G. J., Esq.

\subscriber{Moore}, The Hon. Mrs. \textsc{Montgomery},
Dublin.
\columnbreak

\subscriber{Newman}, The Rev. W. A., D.D., Wolverhampton.

\subscriber{Nichol}, — Esq.

\subscriber{Nicholls}, W. H., Esq.

\subscriber{Nicholson, Alfred}, Esq.

\subscriber{Norbury}, The Right Hon. the Earl of.

\subscriber{Norman}, Rev. R. W., St. Peter’s College,
near Abingdon.

\subscriber{Norwood, Jasper}, Esq., Preston.

\subscriber{Norman}, Mrs., The Rookery, Bromley.
\bigskip

\subscriber{Ollivier}, Mr. \textsc{Robert}.

\subscriber{Osborne}, G. A., Esq.

\subscriber{Ouseley}, The Rev. Sir \textsc{Frederick Arthur}
\subscriber{Gore}, Bart., Professor of Music in the
University of Oxford.

\subscriber{Ould, Edwin}, Esq.
\bigskip

\subscriber{Paget}, R., Esq.

\subscriber{Parker, John William}, Esq.

\subscriber{Paterson}, Mr. R. \textsc{Roy}, Edinburgh.

\subscriber{Pole, S. Chandos}, Esq.

\subscriber{Purdie}, Mr. \textsc{John}, Edinburgh.

\subscriber{Pye, Kellow}, Esq., Wimbledon.
\bigskip

\subscriber{Rawlings}, — Esq., Shrewsbury.

\subscriber{Ravey}, Mr.

\subscriber{Reed, T. German}, Esq.

\subscriber{Richards}, Mr. T.

\subscriber{Richardson}, — Esq., Swindon.

\subscriber{Romer, Frank}, Esq.

\subscriber{Robinson, Joseph}, Esq., Dublin.

\subscriber{Rowe}, Mrs., Plymouth.

\subscriber{Rhodes, Jeremiah}, Esq., Pontefract.

\subscriber{Robb}, Mrs.

\subscriber{Robeck}, The Baron de, Swordlestown.

\subscriber{Rudall, Rose, and Carte}, Messrs.
\bigskip

\subscriber{Stanford, John}, Esq., Dublin.

\subscriber{Salaman, Charles}, Esq.

\subscriber{Sandys}, W., Esq., F.S.A.

\subscriber{Salt, George}, M., Esq., Shrewsbury.

Sacred Harmonic Society (The)

\subscriber{Spark}, W., Esq., Organist, Leeds.

\subscriber{Smith, G. Townshend}, Esq., Organist of the
Cathedral, Hereford.
\end{multicols}
%xix
\pagebreak