%%20
 
\begin{multicols}{2}\small
\setlength{\parindent}{0em}
\setlength{\parskip}{0em}
\raggedright

\subscriber{Spencer}, The Rev. \textsc{Peter}, Temple Ewell,
near Dover.

\subscriber{Simms}, E., Esq., Coventry.

\subscriber{Simms}, Mr. H., Bath.

\subscriber{Smart}, Sir \textsc{George}.

\subscriber{Smith, Montem}, Esq.

\subscriber{Smith, Albert}, Esq.

\subscriber{Smith, Samuel}, Esq., Bradford.

\subscriber{Smith, W. H}., Esq., Sheffield.

\subscriber{Smith}, Mr. \textsc{John}, Leeds.

\subscriber{Smith, George}, Esq.

\subscriber{Singleton}, Rev. R. C., Kingstown, Dublin.

Signet Library, (The) Edinburgh.

\subscriber{Silver, William}, Esq.

\subscriber{Swift, Joseph}, Esq.

\subscriber{Skelton}, G. J., Esq., Hull.

\subscriber{Shea, Alexander}, Esq.

\subscriber{Shelmerdine}, W., Esq., Nottingham.

Shoreham College.

\subscriber{Stanyon, John}, Esq., Leicester.

\subscriber{Stanley}, E. H., Esq.

\subscriber{Stephenson}, W. F., Esq., Organist, Bishopton.

\subscriber{Sloper, Lindsay}, Esq.

\subscriber{Sutton} and \textsc{Potter}, Messrs., Dover.

\subscriber{Stocqueler}, J. H., Esq.

\subscriber{Sykes, Luke}, R., Esq.

\subscriber{Spalding}, S., Esq.

\subscriber{Squire}, F., Esq.

\subscriber{Stokes}, Dr., Dublin.
\bigskip

\subscriber{Travers}, Miss M.T., Hartsbourne, Bushey,
Herts.

\subscriber{Thacker}, A. C., Esq., Peterborough.

\subscriber{Taylor, Hussey}, Esq.

\subscriber{Taylor, Bianchi}, Esq., Bath.

\subscriber{Thackeray}, W., Esq.

\subscriber{Tait}, W., Esq., Melrose.
%\columnbreak

\subscriber{Thomson, James}, Esq., Glasgow.

\subscriber{Thrupp, John}, Esq.

\subscriber{Turner}, Mr., Stockport. 3 \textit{copies}.

\subscriber{Turle}, Miss, Lyme Regis.

\subscriber{Travers}, Miss, Hillingdon.
\bigskip

\subscriber{Vernon}, The Right. Hon. Lord.

\subscriber{Vantini, Townsend}, Esq.
\bigskip

\subscriber{Ward}, The Right Hon. Lord. 2 \textit{copies}.

\subscriber{Walker, George}, Esq., Aberdeen.

\subscriber{Wallerstein}, F. G., Newcastle-on-Tyne.

\subscriber{Walsh}, E., Esq.

\subscriber{Warren}, E., Esq., Organist of Christ
Church, Dover.

\subscriber{Wallace, W. Vincent}, Esq.

\subscriber{Ward}, The Rev. \textsc{Arthur} R., St. John’s
College, Cambridge.

\subscriber{Warren}. Mr. \textsc{John}, Royston.

\subscriber{Weiss}, W. H., Esq.

\subscriber{Werner, Louis}, Esq.

\subscriber{Webster, Alfred}, Esq., Bath.

\subscriber{Williams}, T., Esq., Fairlight, Tunbridge
Wells.

\subscriber{Wighton}, Mr. A. J., Dundee.

\subscriber{Wilson}, M. C., Esq.

\subscriber{Winslow, Forbes}, Esq., M.D., D.C.L.

\subscriber{Winn}, W., Esq.

\subscriber{Winstanley}, J. B., Esq., Bramston House,
near Leicester.

\subscriber{Woolcombe}, The Rev. W. \textsc{Walker}, Manchester.

\subscriber{Wood}, Messrs. J. M., and Co., Glasgow.
2~\textit{copies}.

\subscriber{Wood}, Messrs., Edinburgh. 2~\textit{copies}.

\subscriber{Woodman}, W., Esq., Hobhill, Morpeth.

\subscriber{Wylde, Henry}, Esq., Mus. Doc.
\end{multicols}
%xx
\pagebreak
\setlength{\columnseprule}{0pt}
%===============================================================================


